Towards an Ecological Trait-data Standard

### 2018-05-18 manuscript/manuscript.Rmd
finalize ms

\footnotetext[1]{Corresponding author: florian.dirk.schneider@gmail.com}
\footnotetext[2]{Institute of Linguistics and Literary Studies, Technische Universität Darmstadt, Darmstadt, Germany}
\footnotetext[3]{Senckenberg Biodiversity and Climate Research Centre (BiK-F) Frankfurt am Main, Germany}
\footnotetext[4]{Institute of Plant Sciences, University of Bern, Bern, Switzerland}
\footnotetext[5]{Botanischer Garten und Botanisches Museum Berlin-Dahlem, Freie Universität Berlin, Berlin, Germany}
\footnotetext[6]{Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Jena, Germany}
\footnotetext[7]{Terrestrial Ecology Research Group, Department for Ecology and Ecosystem Management, Center for Life and Food Sciences Weihenstephan, Technische Universität München, Freising, Germany}

@@

@@

As they deal with much larger amounts of data, initiatives that form around natural history museum collections are more concerned with data standardisation. 
With the digitization efforts that are currently undertaken in many museum collections [@vollmar10; @blagoderov12], supported by citizen science crowdsourcing (e.g. www.markmybird.org), data on body measurements are likely to grow exponentially in the near future. 
For example, the VertNet database compiled and harmonized large quantities of vertebrate traitdata with the aim of mobilising measurements from collections [@guralnick16]. The resulting data are published as versioned data tables which are updated as new data sources are integrated.

@@

While trait thesauri define traits for focal groups of organisms, they are not specifying the  format or structure in which trait data should be stored and linked to further standard terminologies, e.g. for taxa. 
To make sense of trait data in the context of more general databases, a consensus definition of trait datasets is inevitable.

Trait data have been defined  by @garnier17 to follow an entity-quality model (EQ), where a trait observation is phrased as 'an entity having a quality'. 
More specifically, a trait dataset contains information on quantitative measurements or qualitative facts (i.e. trait values) describing the physical phenotypic characteristics relating to fitness, behaviour or performance (i.e. traits) observed on a biological entity (i.e. an individual specimen, or parts of an individual specimen) that can be assigned to a biological taxon (i.e. a species or higher-level taxon). 
Expaning on this definition, quantitative measurements are values obtained either by direct morphological, physiological or behavioural observations on single specimens (Fig. 1*a*), by aggregating replicated measurements on multiple entities (Fig. 1*b*) or by estimation the means or ranges for the species as reported in literature or other published sources (e.g. databases, Fig. 1*c*). Qualitative facts are assignments of an entity to a categorical level, e.g. of a behavioural or life-history trait (Fig. 1*d*). 
The entity or observation (i.e. the occurrence) to which the reported measurement or fact applies may differ in organisational scale -- depending on the scientific question -- and could be a sub-sample or bodypart, an individual specimen, an entire species or a higher-level taxon (e.g. a genus).

@@

These relations between a trait observation and an individual specimen as an occurrence of a particular taxon have been formalized in the schema for biological collection records [ABCD Schema; @walls14] and the Darwin Core Standard for biodiversity data [DwC; @wieczorek12]. 
These frameworks specify terms and classes to describe the general structure of biodiversity databases, for example by defining which field is to contain trait values, units, taxon names, variables such as sex or life stage, as well as ancillary information of time and date of observation, as well as methodological details. The terminologies provided by these standards are quite universal and cover most use cases of trait data. 
<!-- TODO FDS: dimensional and star-shema databases, relational databases Kattge et al 2011, Kimball & Ross 2002-->
<!--  DF: am not sure, where to best write this comment, but I think it belongs somewhere in this chapter. The Trait Data Standard is a standard on how to store trait data measurements. In this chapter you mention what databases for trait data there are (usually specific for certain taxa) and what ontologies there for the different traits, however it is missing the section on how trait data measurements have been and can be stored prior to your standard. For instance the ABCD standard has a huge section for MeasurementsAndFacts. And while it certainly is not primarily made to capture trait measurements, it can certainly be used for it (at least on a unit level, combining measurements for higher taxa is not possible). Darwin Core has similar, though not as extensive, fields for measurements, as they were modelled based on ABCD. While I agree that your standard is useful and necessary, it certainly is not the first standardized way to describe trait measurements.  -->

Specifically designed for the context of plant traits, @kattge11 proposed a generic structure for databases that covers most potential use cases of trait-based ecology. This data structure is build around a central data table that contains observations, i.e. a single event of measurements on the same individual plant specimen at the same point in time. This structure puts emphasis on the fact that trait  data are often measured for an analysis of correlations of multiple traits. Identifiers link multiple measurements (qualities) to the same observation (entity), each measurement being well defined by additional standard tables. The observations are also linked to a taxonomy and ancillary descriptors of the observation context, like location or experimental treatment. This structure can be implemented in any software or programming language for relational databases. 
<!-- FDS: enter comments of Jens Kattge here!-->

In a similar vein, the Encyclopedia of Life (EOL) project has proposed the database framework TraitBank [@parr16] for major physiological and life-history traits of all kingdoms, which is to date the most general approach to providing an integrated structure for trait data tables.
The framework employs established terms provided by the DwC, relates trait definitions to specific trait, phenotypic and anatomical ontologies, and maps taxa to global identifiers in taxonomic hierarchies of name service providers  to capture synonyms, misspellings and controversies [@parr16, http://eol.org/info/cp_archives]. The difficulties with keeping taxonomic references intact along with continuous changes in taxonomy consensus are a central challenge of biodiversity data management and are beyond the scope of this review. 
Additional layers of information capture bibliographic reference, multimedia archives and ecological interactions. 
TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive [DwC-A, @robertson09a], a zip-file with annotated text-files that is also preffered for observation data in GBIF [@gbif17,  http://tools.gbif.org/dwca-assistant/]. 
The archive also integrates the general framework for metadata of the Ecological Metadata Language [EML, @knb11].

These proposed standards and the available terms and extensions are responses to a demand from biodiversity data management for more structured input from the research community. 
Hovewer, hardly any of the aforementioned aggregated and harmonized datasets for birds, amphibians, or mammals employs such ontologies or semantic web standards.

One reason for this is most certainly complexity: The data structures are designed for multi-layered, relational databases rather than for static datasets for which a two-dimensional data table might suffice.  
If the traits are assembled for a specific research question with few columns of ancillary information, it may be safer to add this information as extra columns to the two-dimensional data table, thus facilitating an analysis of correlations between traits and co-factors further down the road. This way, if datasets of different source are merged, the related information is readily available without the risk of breaking the reference to an external datasheet.

The other reason is lack of awareness of data providers for trait data standardization: many trait-data providers are not trained in the demands of biodiversity data-management and complying with the apparently non-intuitive data structures is an investment without clear incentive or immediate pay-off. Investing time and energy into research for the best and most sustainable data structure is not affordable for small and intermediate-size research projects.

By filling this gap, data brokering services [e.g. the German Federation on for Biological Data; gfbio.org; @diepenbroek14] or data management systems for scientific projects (e.g. BExIS, NEON Dataportal; Diversity Workbench, diversityworkbench.net; KNB and its open-source database back-end Metacat, https://knb.ecoinformatics.org/) are likely to gain importance. These services simplify and direct the standardized upload of research data and descriptive metadata into reliable and interlinked data infrastructures. One goal of such initiatives is to facilitate data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses. <!-- AG:  EDIT Platform for Cybertaxonomy (cybertaxonomy) -->

@@

Initiatives for standardization (e.g. ontologies and data standards) and platforms for data management (e.g. database and data management platforms) provide great visibility and improve interconnectedness of datasets, but raise relatively high thresholds for data and metadata preparation.  
With low-threshold offers for the hosting of scientific primary data attracting a wealth of heterogeneous trait-datasets, the concern is with data harmonization of these distributed data sources.

The goal must be to integrate these distributed data better into the global biodiversity data-management ecosystem by creating awareness for data standardization on the side of data providers. 
We propose the development of  tools and vocabularies that impose low thresholds and offer high payoff in data visibility and interconnectedness of the published data.

@@

# 2. Proposing an ecological traitdata standard vocabulary

As a response to the observations and conclusions above, we propose a versatile vocabulary that captures the different degrees of measurement detail for multiple use cases of trait-based ecological research. 
Rather than describing a data structure for relational databases, this is intended as a more inclusive terminology, that can be used in simple two-dimensional datasets as well as in the exchange of data between web services in the 'semantic web'.

By using this standard vocabulary, authors can ensure that the description of tables of trait measurements that are uploaded to distributed data repositories like Figshare, Zenodo etc. will be unambiguous and generally applicable. It will facilitate re-use of data for future data aggregation initiatives and data synthesis and ensure accessibility in the long term.

The online reference to the standard vocabulary can be found at https://ecologicaltraitdata.github.io/ETS/.

In designing this vocabulary, we drew on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).  
We paid particular consideration to the work of @kattge11, @kattge11a,  and @garnier17, as well as @parr16 to ensure compatibility of our proposed data structure with major trait databases and existing standards for biodiversity data management. Here, the use of identifiers ('IDs') for the individual measurement observations ('measurementID'), specimens ('occurenceID'), sampling events ('eventID'), or taxa ('taxonID') is key to map two-dimensional data onto the structure of relational databases.

However, the vocabulary proposed is meant as a preliminary draft of a standard nomenclature to be completed and corrected by the wider community of researchers using trait-based approaches in ecology, and may in the future integrate more morphological and evolutionary perspectives as well.
Also, at present we conceptualize trait-datasets to  not include information about a species' interaction partners/associations of an organism, since they are -- to our understanding -- not within the scope of the Entity-Quality Model. However, such network data may easily be combined with trait data via the taxon identifiers or by using other Extensions of the Darwin Core (DwC).
Therefore, we invite researchers to contribute to the next iterations of the standard vocabulary and develop own applications and ontologies that interact with it.

## Why use the vocabulary

The incentives for using a unified vocabulary for trait data might not be obvious for data providers, who are mainly concerned with a particular study site, organism group and research question. We won't discuss here in depth how sharing trait data will improve science on broader questions and larger scales, by providing a data basis for downstream analysis. It appears to be broad consensus that an open biodiversity science is crucial for an evidence based decision making and conservation policy on regional and global scales.

Instead, this section points on the technical aspects and advantages of readying own trait data to link with the 'semantic web' of biodiversity data and knowledge. 
By using a constrained vocabulary with globally accessible definitions of terms, data can be accessed more easily by other researchers and datasets from different sources can be combined into aggregated datasets.
Also it will be easier to develop computational methods and software tools that access and handle the data, based on the standard vocabulary.

It is implicit to this vocabulary to use a two-dimensional observation long-table format for the data, rather than a species-trait matrix (where some of the columns would be trait names which are not included in the vocabulary, but are part of trait thesauri). As this format draws from a defined set of columns,  merging of datasets is  easier. Long-table datasets also purport multiple advantages for data manipulation [e.g. filtering, sub-setting and aggregating data; @wickham14].

Using well-defined identifiers ('IDs') to structure the dataset offers compatibility with the input formats for larger databases that may offer better visibility and impact of the data contribution. 
Identifier fields are key elements to structure the two-dimensional dataset. 
For instance, for occurrence level data, were multiple trait measurements are reported for each individual specimen, the same user-defined entry for 'occurrenceID' would link these measurements across the rows of the dataset. 
Similarly, multivariate measurements, like for instance gas chromatography data or x-y-z data of morphometric landmarks could be linked via a 'measurementID'. 
If literature data are provided, data are probably not structured below the taxon level and a 'taxonID' is the key identifier.
A 'datasetID' allows to report datasets that have been compiled from multiple sources.  
Beyond being just of structural use for the dataset, identifiers are capable of linking own data to consensus taxonomy and trait terminology via Uniform Resource Identifiers (URIs), which point to external ontology services (see above for resources).

However, two-dimensional spreadsheets are limited in the number and complexity of measurements they can contain. For datasets containing multi-layered information on observations, traits, taxa and environmental context, we  recommend to consider relational datatabase structures composed of multiple spreadsheets linked via identifier columns, like for example the generic trait database structure proposed by @kattge11a or the Traitbank structure proposed by @parr16, ideally compiled into Darwin Core Archives [DwC-A, @robertson09a]. The trade-off is data-consistency and readability in a single table *vs.* avoidance of content duplication and redundancy in a relational database. The standard vocabulary proposed here may still be applied within these structured databases.

@@

@@

Column names and, where possible, field entries of a trait datasets published on online repositories should draw from the defined vocabulary.

The core vocabulary lists and defines terms that describe a dataset according to the Entity-Quality model described above [@garnier17]: a trait dataset contains information on quantitative measurements or qualitative facts (i.e. trait values) describing the physical phenotypic characteristics relating to fitness, behaviour or performance (i.e. traits) observed on a biological entity (i.e. an individual specimen, or parts of an individual specimen) that can be assigned to a biological taxon (i.e. a species or higher-level taxon).

Additional terms are provided for metadata and for lists of trait definitions (i.e. trait thesauri, see below).

Beyond the core observations, further information might be available that are related to the individual or specimen or the taxon, or that put the reported fact, measurement or sampling event in a broader observation context  (including geolocation and date information). These information can be useful for future analysis of the causal reasons of trait variation and should always be published along with the core data. For this case, we offer three extensions of the core vocabulary ("taxon", "measurement or fact", and "occurrence") that expand and refine terms of the Darwin Core Extensions (see below). This information may simply be added as extra columns to the core dataset.

For reasons of long-term accessibility, data should not be uploaded in spreadsheet formats (like '.xlsx') but rather in comma-separated text files  ('.csv' or '.txt') that are compatible with all computing platforms and internationalisation settings by applying a unified character encoding (e.g. UTF-8 or ASCII).

In order to ensure traceability, the metadata of any dataset that employs this vocabulary should refer to the specific online version that was used to build the dataset, e.g."Schneider, F.D., Jochum, M.,  LeProvost, G., Penone, C., Ostrowski, A. and Simons, N.K., 2017 Ecological Traitdata Standard v0.6, DOI: 10.5281/zenodo.1048956, URL: https://ecologicaltraitdata.github.io/ETS/v0.6/". In addition to this versioned online reference, the dataset should also cite this paper for an explanation of the rationale.

Wherever referring to individual terms in publications or metadata, please do so via their global identifier, which is hosted by the GFBio Terminology Service [@karam16; https://terminologies.gfbio.org/], which also can be accessed programmatically (i.e. via the Application Programming Interface, API **in preparation!**). Wherever our glossary refines or duplicates existing terms from other ontologies for biological data, like the Glossary of EOL (http://eol.org/info/516) and Darwin Core (http://rs.tdwg.org/dwc/terms/), we indicate this in the fields 'refines' or 'identical', respectively.

Besides this use in published datasets, the standard vocabulary could be imposed in webservices, e.g. APIs that provide direct access to online databases or download tools that pull subsets of data from larger databases.

### Specifications of core terms

To qualify as trait data according to the definition provided above, where each row is the reported measurement or fact for a single observation, the following columns are required at minimum: 
1. a value (column `traitValue`) and -- for numeric values -- a standard unit (`traitUnit`); 
2. a descriptive trait name (`traitName`) that links to a well-defined definition; 
3. the scientific taxon for which the measurement or fact was obtained (`scientificName`). 
For these core values, we highly recommend to use unambiguous and self-explanatory vocabularies for trait names and taxa. 
However, in order to keep the data compatible with own databases or analytical code, it might be necessary to use abbreviations or own identifiers for `scientificName`  and `traitName` instead.

<!--### Relate to global taxon and trait definitions -->

In this case, it is essential to relate the user-defined names to a consensus standard of taxon names as well as a look-up tables of traits. 
This is achieved by adding globally valid Uniform Resource Identifiers (URIs) for taxa (`taxonID`) and trait definitions (`traitID`), complemented by the human-readable verbatim accepted names (`ScientificNameStd` and `traitNameStd`, respectively). For example, referring to GBIF Backbone Terminology, for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; the `traitID` for 'fruit mass' according to TOP Thesaurus of plant traits would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass'.  
By allowing for a double record of both user-specific and standardised entries, we acknowledge the fact that most authors have their own schemes for standardisation which may refer to different scientific community standards  [as practiced in TRY; @kattge11]. This redundancy of data allows for continuity on the authors and data owners side while also ensuring quality checks and comparability for the data user.

<!-- PM: '... failure to provide these URI for taxa will ...' In general some examples of benefits would be good -->

@@

The column containing taxon names is given in parameter 'taxa'. Note that the parameter 'keep' specifies and renames any data that should be maintained in the output. The parameter 'units' is used to specify the input units of measurement.

>>>>

\footnotetext[1]{Corresponding author: florian.dirk.schneider@gmail.com}
\footnotetext[2]{Institute of Linguistics and Literary Studies, Technische Universität Darmstadt, Darmstadt, Germany}
\footnotetext[3]{Senckenberg Biodiversity and Climate Research Centre (BiK-F) Frankfurt am Main, Germany}
\footnotetext[4]{Institute of Plant Sciences, University of Bern, Bern, Switzerland}
\footnotetext[5]{Botanic Garden and Botanical Museum Berlin, Freie Universität Berlin, Berlin, Germany}
\footnotetext[6]{Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Jena, Germany}
\footnotetext[7]{Department for Ecology and Ecosystem Management, Center for Life and Food Sciences Weihenstephan, Technische Universität München, Freising, Germany}

@@

Controlled vocabulary | 

@@

As they deal with much larger amounts of data, initiatives that form around natural history museum collections are more concerned with data standardisation. 
With the digitization efforts that are currently undertaken in many museum collections [@vollmar10; @blagoderov12], supported by citizen science crowdsourcing (e.g. www.markmybird.org), data on body measurements are likely to grow exponentially in the near future. 
For example, the VertNet database compiled and harmonized large quantities of vertebrate trait data with the aim of mobilising measurements from collections [@guralnick16]. The resulting data are published as versioned data tables which are updated as new data sources are integrated.

@@

While trait thesauri and ontologies typically define traits for focal groups of organisms, they do not specify the  format or structure in which trait data should be stored and linked to further standard terminologies, e.g. for taxa. 
To make sense of trait data in the context of more general databases, a consensus definition of trait data is necessary.

Trait data have been defined by @garnier17 to follow an entity-quality model (EQ), where a trait observation is 'an entity having a quality'. 
More specifically, a trait dataset contains information on quantitative *measurements* or qualitative *facts* (i.e. trait values) describing the physical phenotypic characteristics relating to fitness, behaviour or performance (i.e. traits) observed on a biological entity (i.e. an individual specimen, or parts of an individual specimen) that can be assigned to a biological taxon (i.e. a species or higher-level taxon). 
We are expanding on this definition: quantitative measurements are values obtained either by direct morphological, physiological or behavioural observations on single specimens (Fig. 1*a*), by aggregating replicated measurements on multiple entities (Fig. 1*b*) or by estimating the means or ranges for the respective taxon as reported in the literature or other published sources (e.g. databases, Fig. 1*c*). Qualitative facts are assignments of an entity to a categorical level, e.g. of a behavioural or life-history trait (Fig. 1*d*). 
The entity or observation (i.e. the *occurrence*) to which the reported measurement or fact applies may differ in organisational scale -- depending on the scientific question -- and could be a sub-sample or bodypart, an individual specimen, an entire species or a higher-level taxon (e.g. a genus).

@@

These relationships between a trait observation and an individual organism as an occurrence of a particular taxon have been formalised in the schema for biological collection records [ABCD Schema; @walls14] and the Darwin Core Standard for biodiversity data [DwC; @wieczorek12]. For example, the Global Biodiversity Information Facility (GBIF, www.gbif.org) applies these terms.
These frameworks specify terms and classes to describe the general structure of biodiversity databases, for example by defining names for columns that contains measurement values, units, taxon names, variables such as sex or life stage, ancillary information of time and date of observation, and methodological details. The terminologies provided by these standards are quite universal and even cover most use cases of trait data. An entire ecosystem of data standards links to and expands the capacities of DwC [@wieczorek12].  
<!--  DF: am not sure, where to best write this comment, but I think it belongs somewhere in this chapter. The Trait Data Standard is a standard on how to store trait data measurements. In this chapter you mention what databases for trait data there are (usually specific for certain taxa) and what ontologies there for the different traits, however it is missing the section on how trait data measurements have been and can be stored prior to your standard. For instance the ABCD standard has a huge section for MeasurementsAndFacts. And while it certainly is not primarily made to capture trait measurements, it can certainly be used for it (at least on a unit level, combining measurements for higher taxa is not possible). Darwin Core has similar, though not as extensive, fields for measurements, as they were modelled based on ABCD. While I agree that your standard is useful and necessary, it certainly is not the first standardized way to describe trait measurements.  -->

Specifically designed for plant traits, @kattge11 proposed a generic database structure that covers most potential use cases of trait-based ecology. This data structure is built around a central data table that contains observations, i.e. a single event of measurement on the same individual plant specimen at the same point in time. This structure emphasises the fact that multiple trait data are measured on the same individual organisms and used to analyse correlations between these multiple traits. Identifiers link the measurements (qualities) to the same observation (entity), each measurement being well defined by additional standard tables. The observations are also linked to a taxonomy and ancillary descriptors of the observation context, like location or experimental treatment. This structure can be implemented in any relational database management system.

In a similar vein, the Encyclopedia of Life (EOL) project has proposed the database framework TraitBank [@parr16] for major physiological and life-history traits of all kingdoms of life, which is to date the most general approach to providing an integrated structure for trait data.
The framework employs established terms provided by the DwC, relates trait definitions to trait ontologies for phenotypic or anatomical terms, and maps taxa to global identifiers in taxonomic hierarchies of name service providers  to capture synonyms, misspellings and controversies [@parr16, http://eol.org/info/cp_archives]. 
Additional layers of information capture bibliographic reference, multimedia archives and ecological interactions. 
TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive [DwC-A, @robertson09a], a zip-file with annotated text-files that is also preffered for observation data in GBIF [@gbif17,  http://tools.gbif.org/dwca-assistant/]. 
The archive also integrates the general framework for metadata of the Ecological Metadata Language [EML, @knb11]. 
The difficulties with keeping taxonomic references intact along with continuous changes in taxonomy consensus are a central challenge of biodiversity data management and are beyond the scope of this review [@franz16]. Initiatives that aim at providing a stable reference for taxa are for instance the EOL Catalogue of Life [http://www.catalogueoflife.org/, @roskov18], the GBIF Backbone Taxonomy [@gbifsecretariat17], or the EDIT Platform for Cybertaxonomy (https://cybertaxonomy.eu/).

These proposed standards are responses to a demand from biodiversity data managers for more structured input from the research community. 
Hovewer, hardly any of the aforementioned trait datasets for birds, amphibians, or mammals employs such ontologies or semantic web standards. 
One reason for this is most certainly complexity: The data structures are designed for multi-layered, relational databases rather than for standalone datasets for which a two-dimensional data table may suffice. In the eyes of the data-provider, in most cases, ancillary co-factors can be appended as extra columns to the dataset. 
<!--If the traits are assembled for a specific research question with just few ancillary co-factors, it may be safer to append these as columns to the two-dimensional data table, thus facilitating an analysis of correlations between traits and co-factors further down the road. This way, if datasets of different sources are merged, the related information is readily available without the risk of breaking the reference to an external datasheet. -->
The other reason is lack of awareness for the need for trait data standardisation among data providers: many providers are not trained in the demands of biodiversity data-management and complying with the apparently non-intuitive data structures is an investment without clear incentive or immediate pay-off, and hardly affordable for small and intermediate-size research projects.

By filling this gap, data-brokering services [e.g. the German Federation for Biological Data; gfbio.org; @diepenbroek14] or data management systems for scientific projects (e.g. BExIS, http://bexis2.uni-jena.de/; Diversity Workbench, www.diversityworkbench.net; KNB and its open-source database back-end Metacat, https://knb.ecoinformatics.org/) are likely to gain importance. These services simplify and direct the standardised upload of research data and descriptive metadata into reliable and interlinked data infrastructures. One goal of such initiatives is to facilitate data publications and standardisation for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses. <!-- AG:  EDIT Platform for Cybertaxonomy (cybertaxonomy); NEON Dataportal http://data.neonscience.org/home;   -->

@@

Initiatives for standardisation (e.g. ontologies and data standards) and platforms for data management (e.g. database and data management platforms) provide great visibility and improve interconnectedness of datasets, but raise relatively high thresholds for data and metadata preparation. 
Low-threshold repositories offer the hosting of scientific primary data attracting a wealth of heterogeneous trait-datasets, but data harmonisation of these distributed data sets is currently laborious. 
The goal must be to integrate these distributed data better into the global biodiversity data-management ecosystem by creating awareness for data standardisation on the side of data providers. 
We propose the development of  tools and vocabularies that impose low thresholds and offer high payoff in the visibility and interconnectedness of published data.

@@

# An ecological trait-data standard vocabulary

As a response to the challenges outlined above, we propose a versatile vocabulary for trait-based ecological research. 
The aim of the vocabulary is to cover the variety of trait-based approaches and their different degrees of measurement detail. 
Rather than describing a data structure for relational databases, the vocabulary is intended as a more inclusive terminology, that can be used in simple two-dimensional datasets as well as in the exchange of data between web services in the 'semantic web'.  
By using this standard vocabulary, authors can ensure that the description of trait measurements that are uploaded to distributed data repositories will be unambiguous and generally applicable. It will facilitate re-use of data for future data aggregation initiatives and data synthesis and ensure long-term accessibility.

In designing this vocabulary, we drew on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).
We paid particular consideration to the work of @kattge11, @kattge11a, and @garnier17, as well as @parr16 to ensure compatibility of our proposed data structure with major trait databases and existing standards for biodiversity data management. Here, the use of identifiers ('IDs') for the individual measurement observations ('measurementID'), specimens ('occurrenceID'), sampling events ('eventID'), or taxa ('taxonID') is key to map two-dimensional data onto the structure of relational databases. 
Besides being used for the publication of datasets, the standard vocabulary could be imposed in webservices or download tools, e.g. APIs that provide direct access to online databases. 
The vocabulary proposed is intended to form the foundations of a standard nomenclature that can be expanded and corrected by the wider community of researchers using trait-based approaches in ecology.
<!--
## Why use the vocabulary

The incentives for using a unified vocabulary for trait data might not be obvious for data providers, who are mainly concerned with a particular study site, organism group and research question. We won't discuss here how sharing trait data will improve science on broader questions and larger scales, by providing a data basis for downstream analysis. However, it appears to be broad consensus that an open biodiversity science is crucial for an evidence-based decision making and conservation policy on regional and global scales.

Instead, this section points on the technical aspects and advantages of readying own trait data to link with the 'semantic web' of biodiversity data and knowledge. 
By using a constrained vocabulary with globally accessible definitions of terms, data can be accessed more easily by other researchers and datasets from different sources can be combined into aggregated datasets.
Also it will be easier to develop computational methods and software tools that access and handle the data, based on the standard vocabulary. -->

@@

We suggest that any trait dataset that is published on online repositories should draw its column names and field entries from the defined vocabulary where possible. 
The core vocabulary lists and defines terms that describe a dataset according to the Entity-Quality model described above [@garnier17] each entry describes a trait value (i.e. quality) observed on an an individual or population (i.e. entity), of a biological taxon. 
It is implicit to the use of the vocabulary to use a two-dimensional observation long-table format for the data, rather than a species$\times$trait matrix. As the long-table format draws from a defined set of columns, the merging of datasets is  easier. Long-table datasets also purport multiple advantages for data manipulation [e.g. filtering, sub-setting and aggregating data, @wickham14; Fig. 2].

Well-defined identifiers ('IDs') are key elements to structure the datasets and relate them to complementing datasets, if necessary.
For instance, for occurrence level data where multiple trait measurements are reported for each individual specimen, the same user-defined entry for 'occurrenceID' would link several measurements across the rows of the dataset. 
Similarly, multivariate measurements, for instance gas chromatography data or x-y-z data of morphometric landmarks could be linked via a 'measurementID'. 
In literature data, summarized traits are usually given at the taxon level instead of the individual organism (e.g. reported as means or factorials) and a 'taxonID' is the key identifier.
In larger compilations, a 'datasetID' allow to trace data origin to the primary source. 
Beyond being just of structural use for the dataset, identifiers are capable of linking own data to consensus taxonomy and trait terminology via Uniform Resource Identifiers (URIs), which point to external terminology services (see above for resources). 
Two-dimensional spreadsheets are however limited in the number and complexity of co-variates they can contain. As such, for datasets containing multi-layered information on observations, traits, taxa and environmental context, the use of relational datatabase structures may be indicated, like the generic trait database structure proposed by @kattge11a or the TraitBank structure proposed by @parr16. 
The trade-off is user-side readability and handling in a single table *vs.* avoidance of content duplication and redundancy in a relational database. 
The standard vocabulary proposed here may still be applied to describe columns within the individual data tables of relational databases.

@@

For reasons of long-term accessibility, data should not be uploaded in proprietary spreadsheet formats (like '.xlsx') but rather in comma-separated text files  ('.csv' or '.txt') that are compatible with all computing platforms and internationalisation settings by applying a unified character encoding (e.g. UTF-8 or ASCII).

In order to ensure traceability, the metadata of any dataset that employs this vocabulary should refer to the specific online version that was used to build the dataset, e.g."Schneider, F.D., Jochum, M.,  LeProvost, G., Penone, C., Ostrowski, A. and Simons, N.K., 2017 Ecological Traitdata Standard v0.6, DOI: 10.5281/zenodo.1048956, URL: https://ecologicaltraitdata.github.io/ETS/v0.6/". In addition to this versioned online reference, the dataset should also cite this paper for an explanation of the rationale.
Wherever referring to individual terms of the vocabulary in publications or metadata,this should be done via their global identifiers, which will be hosted by the GFBio Terminology Service [@karam16, https://terminologies.gfbio.org/] and can be accessed programmatically (i.e. via the API; **in preparation!**). 
Wherever our glossary refines or duplicates existing terms from other ontologies for biological data, like the Glossary of EOL (http://eol.org/info/516) and Darwin Core (http://rs.tdwg.org/dwc/terms/), we indicate this in the fields 'refines' or 'identical', respectively.

## Terms of the standard vocabulary

The online reference to the standard vocabulary can be found at https://ecologicaltraitdata.github.io/ETS/.
The core terms describe minimal trait data according to the Entity-Quality model. 
Beyond these core observations, further information might be available that are related to the individual or specimen or the taxon, or that put the reported fact, measurement or sampling event in a broader observation context  (including geolocation and date information). These information can be useful for future analysis of the causal reasons of trait variation and should always be published along with the core data. For this case, we offer three extensions of the core vocabulary ("Taxon", "Measurement or Fact", and "Occurrence") that expand and refine terms of the Darwin Core Extensions (see below). This information may simply be added as extra columns to the core dataset.
Additional terms are provided for metadata and for lists of trait definitions (i.e. trait thesauri, see below). 
The scope of the vocabulary may not yet cover all aspects morphological and evolutionary perspectives. 
Also, information about interactions between species are not within the scope of the Entity-Quality Model, but may easily be combined with trait data by using other Extensions of the Darwin Core.
Therefore, we invite researchers to contribute to the next iterations of the standard vocabulary and develop own applications and ontologies that interact with it.

### Specification of core terms

To qualify as trait data according to the definition provided above, where each row is the reported measurement or fact for a single observation, the following columns are required at minimum: 
1. a value (column `traitValue`) and -- for numeric values -- a standard unit (`traitUnit`); 
2. a descriptive trait name (`traitName`) that links to a well-defined definition; 
3. the scientific taxon for which the measurement or fact was obtained (`scientificName`). 
For these core values, unambiguous and self-explanatory vocabularies for trait names and taxa are recommended. 
However, to ensure compatibility with existing databases or analytical code, it might be necessary to use abbreviations or user-specific identifiers for `scientificName`  and `traitName` instead. 
<!--### Relate to global taxon and trait definitions -->
In this case, it is essential to relate the user-defined names to a consensus standard of taxon names as well as a look-up table of traits. 
This is achieved by adding globally valid Uniform Resource Identifiers (URIs) for taxon (`taxonID`) and trait definitions (`traitID`), complemented by the human-readable verbatim accepted names (`ScientificNameStd` and `traitNameStd`, respectively). 
For example, referring to GBIF Backbone Terminology, for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; the `traitID` for 'fruit mass' according to TOP Thesaurus of plant traits would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass'.  
By allowing for a double record of both user-specific and standardised entries, we acknowledge the fact that most authors have their own schemes for standardisation which may refer to different scientific community standards  [as practiced in TRY; @kattge11]. This redundancy of data allows for continuity for data owners while also ensuring quality checks and comparability for the data user.

Beyond measurement units or higher taxon information, further information might be available that may not be core data, but are related to the individual or specimen, or to the reported fact, measurement or sampling event. 
The data standard provides three extensions of the vocabulary that should be used to describe this information:

- The `Taxon` extension (http://ecologicaltraitdata.github.io/ETS/#extension-taxon) provides further terms for specifying the taxonomic resolution of the observation and to ensure the correct reference in case of synonyms and homonyms.
- The `MeasurementOrFact` extension (https://ecologicaltraitdata.github.io/ETS/#extension-measurement-or-fact) provides terms to describe information at the level of single measurements or reported facts, such as the original literature from where the value is cited, the method of measurement or statistical method of aggregation. It provides important information that allows for the tracking of potential sources of noise or bias in measured data (e.g. variation in measurement method) or aggregated values (e.g. statistical method applied), as well as the source of reported facts (e.g. literature source or expert reference).   
- The `Occurrence` extension (https://ecologicaltraitdata.github.io/ETS/#extension-occurrence) contains vocabulary to describe information on the level of individual specimens, such as sex, life stage or age. This also includes the method of sampling and preservation, as well as date and geographical location, which provides an important resource to analyse trait variation due to differences in space and time.

Many terms of these extensions refine or copy terms of the DwC and their own [Taxon](http://rs.tdwg.org/dwc/terms/index.htm#Taxon),  [MeasurementOrFact](http://rs.tdwg.org/dwc/terms/index.htm#measureindex) and [Occurrence](http://rs.tdwg.org/dwc/terms/index.htm#Occurrence) extensions and EOL TraitBank's use of those terms (http://eol.org/info/structured_data_archives).
<!-- If no taxonID has been provided, in order to be unambiguous about the taxon referred to by the trait observation, it is highly recommended to keep an additional reference to higher taxonomic hierarchies (using columns `kingdom`, `order` etc.) to avoid taxonomic homonyms, and provide the authority (???) and key of reference (???).  
Furthermore, it is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for traits recorded at the family or genus level. This information can then be used for trait inference at higher taxonomic resolution. <!-- using hierarchical probabilistic matrix factorisation [@schrodt15]. -->
These additional layers of information can either be added as extra columns to the core dataset or kept in separate data sheets (published separately or as part of a Darwin Core Archive), thus avoiding redundancy and duplication of content. A unique identifier would link to these other datasheets, encoding each individual occurrence of a species (`occurrenceID`), single measurements or reported facts (`measurementID`), locations of sampling (`locationID`) and sampling campaigns (`eventID`). 
Some data-types may directly refer to existing global identifiers for occurrence IDs, e.g. a GBIF URI or a museum collection code references the precise specimen from which the measurement was taken.

@@

\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth, angle=0]{figures/fig3.png}
\caption{Process chart of the functions provided within the R package 'traitdataform' to apply the standard vocabulary to any trait-data table.}
